Analysis-ready VCF at Biobank scale using Zarr

### 2025-01-21 paper.tex

Another advantage of Zarr is the availabilty of several 
independent implementations across different programming languages
(see Methods). Figure~\ref{fig-whole-matrix-compute-zarr-versions}
compares the CPU time required to perform the computations
of Figure~\ref{fig-whole-matrix-compute} 
for af-dist implementations using the Zarr-Python~\cite{zarrpython},
JZarr~\cite{jzarr},
and TensorStore~\citep{tensorstore} (C++ and Python) libraries.
The program using the Tensorstore C++ API is fastest, requiring 
about 20\% less CPU time than the version using Zarr-Python and Numba
used in our other benchmarks.

% num_samples  num_sites       tool  user_time  sys_time     wall_time
% storage  total_time
% 22      1000000    7254858       zarr   16709.55     38.14  16910.057917
% hdd    4.652136
% 20      1000000    7254858     ts_cpp   13233.50     92.81  13573.456810
% hdd    3.701753
% 21      1000000    7254858  zarr_java   43780.50    245.28  44134.462377
% hdd   12.229383
% 23      1000000    7254858      ts_py   18027.43    642.40  19163.174182
% hdd    5.186064
% Cpp / Zarr python [0.79197226]

>>>>

The discrepancy between elapsed time and total CPU time for Zarr 
benchmarks in
Figs~\ref{fig-whole-matrix-compute}--~\ref{fig-column-extract}
and~\ref{fig-whole-matrix-decode} are mostly due to the overhead
of accessing a large number of small chunk files and the reduced
effectiveness of file-system read-ahead caching.
While having many independent chunks maps 
naturally to high-performance parallel access patterns
in the cloud store setting~\citep{durner2023exploiting} 
(see also the section on accessing Genomics England data stored on S3), 
such file access patterns can be problematic in the 
classical file-system setting.
The Zarr Zip store (which is simply a Zip archive of the standard directory 
hierarchy) provides a useful alternative here and can 
significantly reduce overall processing time. To illustrate this
we repeated the benchmarks in Fig~\ref{fig-whole-matrix-decode} in
which we measure the time required to decode the entire genotype
matrix (thus serving as a lower-bound for any subsequent computations).
In the case of $10^6$ samples, the Zip file requires slightly more 
% Full data table above ^
% 58         48129895      1000000    7254858                zarr    23.752762
% 59         48129895      1000000    7254858            zarr.zip    24.240338
storage space than the directory hierarchy (23.75GiB and 24.24G
for file system and Zip file, respectively)
and while total 
CPU time is similar (193m vs 187m), the elapsed wall clock
time for the Zip file version is significantly less 
(258m vs 188m).
        % tool  user_time  sys_time    cpu_time   wall_time
% 11      zarr   11383.82    191.01  192.913833  258.598778
% 29  zarr.zip   11198.37     43.13  187.358333  188.659252

Another advantage of Zarr is the availabilty of several 
independent implementations across different programming languages
(see Methods). Fig~\ref{fig-whole-matrix-compute-zarr-versions}
compares the CPU time required to perform the computations
of Fig~\ref{fig-whole-matrix-compute} 
for af-dist implementations using the Zarr-Python~\cite{zarrpython},
JZarr~\cite{jzarr},
and TensorStore~\citep{tensorstore} (C++ and Python) libraries.
The program using the Tensorstore C++ API is fastest, requiring 
about 20\% less CPU time than the version using Zarr-Python and Numba
used in our other benchmarks.

% num_samples  num_sites       tool  user_time  sys_time     wall_time
% storage  total_time
% 22      1000000    7254858       zarr   16709.55     38.14  16910.057917 hdd    4.652136
% 20      1000000    7254858     ts_cpp   13233.50     92.81  13573.456810 hdd    3.701753
% 21      1000000    7254858  zarr_java   43780.50    245.28  44134.462377 hdd   12.229383
% 23      1000000    7254858      ts_py   18027.43    642.40  19163.174182 hdd    5.186064
% Cpp / Zarr python [0.79197226]
